Parametricity for Primitive Nested Types and GADTs

### 2021-06-25 21-lmcs/21-lmcs.tex
third LMCS revisions

{\small
\[\begin{array}{l}
\mathtt{data\; List \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{nil\;:\; List\;A}\\
\hspace*{0.4in}\mathtt{cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
\end{array}\]}

@@

Over time, there has been a notable trend toward data types whose
non-regular indexing can capture invariants and other sophisticated
properties that can be used for program verification and other
applications.  A simple example of such a type is given by Bird and
Meertens'~\cite{bm98} prototypical nested type
\[\begin{array}{l}
\mathtt{data\; PTree\;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{pleaf\;:\;A \rightarrow PTree\;A}\\
\hspace*{0.4in}\mathtt{pnode\;:\;PTree\;(A \times A) \rightarrow PTree\;A}
\end{array}\]
\noindent
of perfect trees, which can be thought of as constraining lists to
have lengths that are powers of 2.  The above code makes clear that
perfect trees at index type $\mathtt{A}$ are defined in terms of
perfect trees at index type $\mathtt{A \times A}$. This is typical of
nested types, one type instance of which can depend on others, so that
the entire family of types must actually be defined at once. A nested
type thus defines not a family of inductive types, but rather an {\em
  inductive family of types}.  Nested types include simple nested
types, like perfect trees, none of whose recursive occurrences occur
below another type constructor; ``deep'' nested types~\cite{jp20},
such as the nested type {\small
\[\begin{array}{l}
\mathtt{data\; PForest\;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{fempty\;:\;PForest\;A}\\
\hspace*{0.4in}\mathtt{fnode\;:\; A \rightarrow PTree\;(PForest\;A) \to
PForest\;A}
\end{array}\]}
\hspace{-0.04in}of perfect forests, whose recursive occurrences appear
below type constructors for other nested types; and truly nested
types, such as the nested type {\small
\[\begin{array}{l}
\mathtt{data\; Bush\;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{bnil\;:\; Bush\;A}\\
\hspace*{0.4in}\mathtt{bcons\;:\;A \rightarrow Bush\;(Bush \; A)
  \rightarrow Bush\;A} 
\end{array}\]}
\hspace{-0.04in}of bushes, whose recursive occurrences appear below
their own type constructors.

@@

\noindent
{\bf Related Work}\/ There is a long line of work on categorical
models of parametricity for System F; see,
e.g.,~\cite{bfss90,bm05,dr04,gjfor15,has94,jac99,mr92,rr94}.  To our
knowledge, all such models treat ADTs via their Church encodings,
verifying in the just-constructed parametric model that each Church
encoding is isomorphic to the data type it encodes.  The present paper
draws on this rich tradition of categorical models of parametricity
for System F, but modifies them to treat nested types (and therefore
ADTs) as primitive data types.

@@

If $\Gamma$ is a finite subset of\, $\tvars$, $\Phi$ is a finite
subset of\, $\fvars$, $\overline{\alpha}$ is a finite subset of\,
$\fvars^0$ disjoint from $\Phi$, and $\phi^k \in \fvars^k \setminus
\Phi$, then the set $\mcF$ of well-formed types is given in
Definition~\ref{def:wftypes}. The notation there entails that an
application $F F_1...F_k$ is allowed only when $F$ is a type variable
of arity $k$, or $F$ is a subexpression of the form $\mu
\phi^{k}.\lambda \alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity
$k$ then $F$ must be applied to exactly $k$ arguments.  Accordingly,
an overbar indicates a sequence of types whose length matches the
arity of the type applied to it.  Requiring that types are always in
such \emph{$\eta$-long normal form} avoids having to consider
$\beta$-conversion at the level of types. In a type
$\Nat^{\ol{\alpha}}F\,G$, the $\Nat$ operator binds all occurrences of
the variables in $\ol{\alpha}$ in $F$ and $G$; intuitively,
$\Nat^{\ol{\alpha}}F\,G$ represents the type of a natural
transformation in $\ol{\alpha}$ from the functor $F$ to the functor
$G$.  In a subexpression $\mu \phi^k.\lambda \ol{\alpha}.F$, the $\mu$
operator binds all occurrences of the variable $\phi$, and the
$\lambda$ operator binds all occurrences of the variables in
$\ol{\alpha}$, in the body $F$.

@@

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[3ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
\BIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F)\,\ol{G}$}  
\DisplayProof
\\[4ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof
\end{array}\]
\end{defi}
We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
Definition~\ref{def:wftypes} ensures that the expected weakening rules
for well-formed types hold, i.e., if $\Gamma; \Phi \vdash F$ is
well-formed, then both $\Gamma, \ol{\psi}; \Phi \vdash F$ and $\Gamma;
\Phi, \ol{\phi} \vdash F$ are also well-formed. Note, however, that
weakening does not change the contexts in which types can be formed.
For example, the functorial context in which a type $\Nat^\emptyset F
\,G$ is formed must be empty, although it can be weakened by a
functorial variable $\beta$ {\em after} the type is formed; by
contrast, we cannot {\em first} weaken the functorial contexts of $F$
and $G$ to $\Gamma; \beta \vdash F$ and $\Gamma; \beta \vdash G$, and
{\em then} form $\Nat^\emptyset F \,G$ in the weakened context
$\Gamma; \beta$.  If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset
\vdash G$, then our rules allow formation of the type
$\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$, which represents the
System F type $\Gamma \vdash F \to G$. Similarly, if
$\Gamma;\ol{\alpha} \vdash F$, then our rules allow formation of the
type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which
represents the System F type $\Gamma; \emptyset \vdash \forall
\ol\alpha . F$. However, some System F types, such as $\forall
\alpha. (\alpha \to \alpha) \to \alpha$, are not representable in our
calculus. Note that, in the rule for $\mu$-types, no variables in
$\Phi$ are applied to types involving variables from $\ol\alpha$ in
the body $F$ of $\mu \phi. \lambda \ol\alpha.F$. This will be critical
to proving the Identity Extension Lemma for our calculus.

Definition~\ref{def:wftypes} allows the formation of all of the types
from Section~\ref{sec:intro}:
\[\begin{array}{lll}
\mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
\beta\; = \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
\beta)\,\alpha\\ 
\mathit{PTree}\,\alpha & = & (\mu \phi. \lambda \beta.\,\beta +
\phi\,(\beta \times \beta))\,\alpha\\
\mathit{Forest}\,\alpha & = & (\mu \phi. \lambda \beta. \,\onet +
\beta \times \mathit{PTree}\,(\phi \beta))\,\alpha\\ 
\mathit{Bush}\,\alpha & = & (\mu \phi.\lambda \beta. \,\onet + \beta
\times \phi\,(\phi\beta))\,\alpha 
\end{array}\]

@@

If $\Gamma; \Phi \vdash F$ is a type, if $\Gamma$ and $\Phi$ contain
only type variables of arity $0$, and if $k=0$ for every occurrence of
$\phi^k$ bound by $\mu$ in $F$, then we say that $F$ is {\em
  first-order}; otherwise we say that $F$ is {\em
  second-order}. Substitution for first-order types is the usual
capture-avoiding textual substitution. We write $F[\alpha := \sigma]$
for the result of substituting $\sigma$ for $\alpha$ in $F$, and
$F[\alpha_1 := F_1,...,\alpha_k := F_k]$, or $F[\ol{\alpha := F}]$
when convenient, for $F[\alpha_1 := F_1][\alpha_2 := F_2,...,\alpha_k
  := F_k]$. The operation of {\em second-order type substitution along
  $\ol\alpha$} is given in Definition~\ref{def:second-order-subst},
where we adopt a similar notational convention for vectors of types.
Of course, $(\cdot)[\phi^0 :=_\emptyset F]$ coincides with first-order
substitution. We omit $\ol\alpha$ when convenient, but note that it
is not correct to substitute along non-functorial variables.

\begin{defi}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$ and $\Gamma;\Phi, \ol{\alpha}
\vdash F$ with $|\ol\alpha| = k$, then the operation $H[\phi :=_{\ol
    \alpha} F]$ of {\em second-order type substitution along
  $\ol\alpha$} is defined by induction on $H$ as follows:
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, G \,K\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}. G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G)\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{defi}
\noindent
The idea is that the arguments to $\phi$ get substituted for the
variables in $\ol\alpha$ in each $F$ replacing an occurrence of
$\phi$. It is not hard to see that $\Gamma;\Phi \vdash H[\phi
  :=_{\ol{\alpha}} F]$. {\color{red} REMOVE? In addition, if
  \,$\Gamma, \phi^k; \Phi \vdash H$,\, $\Gamma; \ol\psi,\ol{\alpha}
  \vdash F$, $|\ol\alpha| = k$, and $\Phi \cap \ol\psi = \emptyset$,
  then $\Gamma,\ol\psi';\Phi \vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi
        :== \psi'}]]$.}

@@

\[\mathit{pnode}\,(\mathit{pnode}\,
(\mathit{pleaf}\,1,\mathit{pleaf}\,2))\,(\mathit{pnode}\,
(\mathit{pleaf}\,3,\mathit{pleaf}\,4) )\]

@@

\[\mathit{pnode}\,(\mathit{pnode}\,
(\mathit{pleaf}\,4,\mathit{pleaf}\,3))\,(\mathit{pnode}\,
(\mathit{pleaf}\,2,\mathit{pleaf}\,1) )\] i.e., maps $((1,2),(3,4))$

@@

  better name for this operation than reverse.}

@@

Let $ \vdash g : \Nat^\alpha \,\onet\,\alpha$, let $G^\set =
\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$ and $G^\rel =
\relsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$.  By
Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho)) =
G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
instantiations of $a$ and $b$ gives that $(G^\set,G^\set) \;= \;
(G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
  \Nat^\alpha \,\onet\,\alpha}\rho$ $ = \{\eta : K_1 \Rightarrow
\id\}$ $ = \{(\eta_1 : K_1 \Rightarrow \id, \eta_2 : K_1 \Rightarrow
\id)\}$ That is, $G^\set$ is a natural transformation from the
constantly $1$-valued functor to the identity functor in $\set$. In
particular, for every $S : \set$, $G^\set_S : 1 \to S$. Note, however,
that if $S = \emptyset$, then there can be no such morphism, so no
such natural transformation, and thus no term $\vdash g : \Nat^\alpha
\onet \,\alpha$, can exist.  That is, our calculus admits no
non-terminating terms, i.e., terms with the closed type $\Nat^\alpha
\onet \,\alpha$ of the polymorphic bottom.

@@

Let $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$, \/$G^\set =
\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$, and $G^\rel =
\relsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  By
Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho)) =
G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
instantiations of $a$ and $b$ gives that $(G^\set, G^\set) \; = \;
(G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
  \Nat^\alpha \,\alpha\,\alpha}\rho$ $ = \{\eta : \id \Rightarrow
\id\}$ $ = \{(\eta_1 : \id \Rightarrow \id, \eta_2 : \id \Rightarrow
\id)\}$ That is, $G^\set$ is a natural transformation from the
identity functor on $\set$ to itself. Now let $S$ be any set.  If $S =
\emptyset$, then there is exactly one morphism $\id_S: S \to S$, so
$G^\set_S : S \to S$ must be $\id_S$. If $S \not = \emptyset$, if $a$
is any element of $S$, and if $K_a :S \to S$ is the constantly
$a$-valued morphism on $S$, then instantiating the naturality square
implied by the above equality gives that $G^\set_S \circ K_a = K_a
\circ G^\set_S$, i.e., $G^\set_S \, a = a$, i.e., $G^\set_S = \id_S$.
Putting these two cases together we have that for every $S : \set$,
$G^\set_S = \id_S$, i.e., $G^\set$ is the identity natural
transformation for the identity functor on $\set$. So every closed
term $g$ of closed type $\Nat^\alpha\alpha\,\alpha$ always denotes the
identity natural transformation for the identity functor on $\set$,
i.e., every closed term $g$ of type $\Nat^\alpha\alpha\,\alpha$
denotes the polymorphic identity function.

@@

As in~\cite{jg10}, replacing $\onet$ with any type $\emptyset;\alpha
\vdash C$ generalizes Theorem~\ref{thm:short-cut-nested} to deliver a
more general a free theorem whose conclusion is $\mathit{fold}_{H}\, B
\; \circ \; G\; \mu H \; \mathit{in}_{H} = G
\,\setsem{\emptyset;\alpha \vdash K}\, B$.

@@

@@

\noindent
respectively. Note that both of these rules make essential use of our
generalized $\Nat$-types.

>>>>

\[\begin{array}{l}
\mathtt{data\; List \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{nil\;:\; List\;A}\\
\hspace*{0.4in}\mathtt{cons\;:\;A \rightarrow List\;A \rightarrow List\;A}
\end{array}\]

@@

Over time, there has been a notable trend toward data types whose
non-regular indexing can capture invariants and other sophisticated
properties that can be used for program verification and other
applications.  A simple example of such a type is given by the type 
\[\begin{array}{l}
\mathtt{data\; PTree\;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{pleaf\;:\;A \rightarrow PTree\;A}\\
\hspace*{0.4in}\mathtt{pnode\;:\;PTree\;(A \times A) \rightarrow PTree\;A}
\end{array}\]
\noindent
of perfect trees, which can be thought of as constraining lists to
have lengths that are powers of 2.  A similar data type
$\mathtt{Nest}$ is given in the canonical paper~\cite{bm98} on nested
types. The above code makes clear that perfect trees at index type
$\mathtt{A}$ are defined in terms of perfect trees at index type
$\mathtt{A \times A}$. This is typical of nested types, one type
instance of which can depend on others, so that the entire family of
types must actually be defined at once. A nested type thus defines not
a family of inductive types, but rather an {\em inductive family of
  types}.  Nested types include simple nested types, like perfect
trees, none of whose recursive occurrences occur below another type
constructor; ``deep'' nested types~\cite{jp20}, such as the nested
type
\[\begin{array}{l}
\mathtt{data\; PForest\;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{fempty\;:\;PForest\;A}\\
\hspace*{0.4in}\mathtt{fnode\;:\; A \rightarrow PTree\;(PForest\;A) \to
PForest\;A}
\end{array}\]
\hspace{-0.04in}of perfect forests, whose recursive occurrences appear
below type constructors for other nested types; and truly nested
types, such as the nested type {\small
\[\begin{array}{l}
\mathtt{data\; Bush\;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{bnil\;:\; Bush\;A}\\
\hspace*{0.4in}\mathtt{bcons\;:\;A \rightarrow Bush\;(Bush \; A)
  \rightarrow Bush\;A} 
\end{array}\]
\hspace{-0.04in}of bushes, whose recursive occurrences appear below
their own type constructors.

@@

\noindent
{\bf Related Work}\/ There is a long line of work on categorical
models of parametricity for System F; see,
e.g.,~\cite{bfss90,bm05,dr04,gjfor15,has94,jac99,mr92,rr94}.  To our
knowledge, all such models treat ADTs via their Church encodings,
verifying in the just-constructed parametric model that each Church
encoding is a solution for the fixpoint equation associated with its
corresponding data type. The present paper draws on this rich
tradition of categorical models of parametricity for System F, but
modifies them to treat nested types (and therefore ADTs) as primitive
data types.

@@

If $\Gamma$ is a finite subset of\, $\tvars$, $\Phi$ is a finite
subset of\, $\fvars$, $\overline{\alpha}$ is a finite subset of\,
$\fvars^0$ disjoint from $\Phi$, and $\phi^k \in \fvars^k \setminus
\Phi$, then the set $\mcF$ of well-formed types is given in
Definition~\ref{def:wftypes}. The notation there entails that an
application $F F_1...F_k$ is allowed only when $F$ is a type variable
of arity $k$, or $F$ is a subexpression of the form $\mu
\phi^{k}.\lambda \alpha_1...\alpha_k.F'$. Accordingly, an overbar
indicates a sequence of types whose length matches the arity of the
type applied to it.  Requiring that types are always in such
\emph{$\eta$-long normal form} avoids having to consider
$\beta$-conversion at the level of types. In a type
$\Nat^{\ol{\alpha}}F\,G$, the $\Nat$ operator binds all occurrences of
the variables in $\ol{\alpha}$ in $F$ and $G$; intuitively,
$\Nat^{\ol{\alpha}}F\,G$ represents the type of a natural
transformation in $\ol{\alpha}$ from the functor $F$ to the functor
$G$.  In a subexpression $\mu \phi^k.\lambda \ol{\alpha}.F$, the $\mu$
operator binds all occurrences of the variable $\phi$, and the
$\lambda$ operator binds all occurrences of the variables in
$\ol{\alpha}$, in the body $F$.

@@

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[3ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
\BIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F)\,\ol{G}$}  
\DisplayProof
\\[4ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof
\end{array}\]
\end{defi}
We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
Definition~\ref{def:wftypes} ensures that the expected weakening rules
for well-formed types hold, i.e., if $\Gamma; \Phi \vdash F$ is
well-formed, then both $\Gamma, \ol{\psi}; \Phi \vdash F$ and $\Gamma;
\Phi, \ol{\phi} \vdash F$ are also well-formed. Note, however, that
weakening does not change the contexts in which types can be formed.
For example, the functorial context in which a type $\Nat^\emptyset F
\,G$ is formed must be empty, although it can be weakened by a
functorial variable $\beta$ {\em after} the type is formed; by
contrast, we cannot {\em first} weaken the functorial contexts of $F$
and $G$ to $\Gamma; \beta \vdash F$ and $\Gamma; \beta \vdash G$, and
{\em then} form $\Nat^\emptyset F \,G$ in the weakened context
$\Gamma; \beta$.  If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset
\vdash G$, then our rules allow formation of the type
$\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$, which represents the
System F type $\Gamma \vdash F \to G$. Similarly, if
$\Gamma;\ol{\alpha} \vdash F$, then our rules allow formation of the
type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which
represents the System F type $\Gamma; \emptyset \vdash \forall
\ol\alpha . F$. However, some System F types, such as $\forall
\alpha. (\alpha \to \alpha) \to \alpha$, are not representable in our
calculus. Note that, in the rule for $\mu$-types, no variables in
$\Phi$ appear in the body $F$ of $\mu \phi. \lambda \ol\alpha.F$. This
will be critical to proving the Identity Extension Lemma for our
calculus.

Definition~\ref{def:wftypes} allows the formation of all of the types
from Section~\ref{sec:intro}:
\[\begin{array}{lll}
\mathit{List}\, \alpha & = & \mu \beta. \,\onet + \alpha \times
\beta\; \mbox{ or } \; (\mu \phi. \lambda \beta.\,\onet + \beta \times \phi
\beta)\,\alpha\\ 
\mathit{PTree}\,\alpha & = & (\mu \phi. \lambda \beta.\,\beta +
\phi\,(\beta \times \beta))\,\alpha\\
\mathit{Forest}\,\alpha & = & (\mu \phi. \lambda \beta. \,\onet +
\beta \times \mathit{PTree}\,(\phi \beta))\,\alpha\\ 
\mathit{Bush}\,\alpha & = & (\mu \phi.\lambda \beta. \,\onet + \beta
\times \phi\,(\phi\beta))\,\alpha 
\end{array}\]

@@

If $\Gamma; \Phi \vdash F$ is a type, if $\Gamma$ and $\Phi$ contain
only type variables of arity $0$, and if $k=0$ for every occurrence of
$\phi^k$ bound by $\mu$ in $F$, then we say that $F$ is {\em
  first-order}; otherwise we say that $F$ is {\em
  second-order}. Substitution for first-order types is the usual
capture-avoiding textual substitution. We write $H[\alpha := F]$
for the result of substituting $F$ for $\alpha$ in $H$, and
$H[\alpha_1 := F_1,...,\alpha_k := F_k]$, or $H[\ol{\alpha := F}]$
when convenient, for $H[\alpha_1 := F_1][\alpha_2 := F_2,...,\alpha_k
  := F_k]$. The operation of {\em second-order type substitution along
  $\ol\alpha$} is given in Definition~\ref{def:second-order-subst},
where we adopt a similar notational convention for vectors of types.
Of course, $(\cdot)[\phi^0 :=_\emptyset F]$ coincides with first-order
substitution. We omit $\ol\alpha$ when convenient, but note that it
is not correct to substitute along non-functorial variables.

\begin{defi}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$ and $\Gamma;\Phi, \ol{\alpha}
\vdash F$ with $|\ol\alpha| = k$, then the operation $H[\phi :=_{\ol
    \alpha} F]$ of {\em second-order type substitution along
  $\ol\alpha$} is defined by induction on $H$ as follows:
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, G \,K\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}. G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G)\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{defi}
\noindent
The idea is that the arguments to $\phi$ get substituted for the
variables in $\ol\alpha$ in each $F$ replacing an occurrence of
$\phi$. It is not hard to see that $\Gamma;\Phi \vdash H[\phi
  :=_{\ol{\alpha}} F]$. {\color{red} REMOVE? In addition, if
  \,$\Gamma, \phi^k; \Phi \vdash H$,\, $\Gamma; \ol\psi,\ol{\alpha}
  \vdash F$, $|\ol\alpha| = k$, and $\Phi \cap \ol\psi = \emptyset$,
  then $\Gamma,\ol\psi';\Phi \vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi
        :== \psi'}]]$. Ask Daniel.}

@@

\[\mathit{pnode}\,(\mathit{pnode}\,(\mathit{pleaf}\,
((1, 2), (3, 4))))\]

@@

\[\mathit{pnode}\,(\mathit{pnode}\,(\mathit{pleaf}\,
((4, 3), (2, 1))))\]
i.e., maps $((1,2),(3,4))$

@@

  better name for this operation than reverse. Ask Daniel.}

@@

Let $ \vdash g : \Nat^\alpha \,\onet\,\alpha$, let $G^\set =
\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$ and $G^\rel =
\relsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$.  By
Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho)) =
G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
instantiations of $a$ and $b$ gives that $(G^\set,G^\set) \;= \;
(G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
  \Nat^\alpha \,\onet\,\alpha}\rho$ $ = \{\eta : K_1 \Rightarrow
\id\}$ $ = \{(\eta_1 : K_1 \Rightarrow \id, \eta_2 : K_1 \Rightarrow
\id)\}$ That is, $G^\set$ is a natural transformation from the
constantly $1$-valued functor to the identity functor in $\set$. In
particular, for every $S : \set$, $G^\set_S : 1 \to S$. Note, however,
that if $S = \emptyset$, then there can be no such morphism, so no
such natural transformation, and thus no term $\vdash g : \Nat^\alpha
\onet \,\alpha$, can exist.  That is, our calculus admits no
(non-terminating) terms with the closed type $\Nat^\alpha \onet
\,\alpha$ of the polymorphic bottom.

@@

Let $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$, \/$G^\set =
\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$, and $G^\rel =
\relsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  By
Theorem~\ref{thm:at-gen}, $(G^\set(\pi_1\rho),G^\set(\pi_2\rho)) =
G^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
instantiations of $a$ and $b$ gives that $(G^\set, G^\set) \; = \;
(G^\set(\pi_1 \rho), G^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
  \Nat^\alpha \,\alpha\,\alpha}\rho$ $ = \{\eta : \id \Rightarrow
\id\}$ $ = \{(\eta_1 : \id \Rightarrow \id, \eta_2 : \id \Rightarrow
\id)\}$ That is, $G^\set$ is a natural transformation from the
identity functor on $\set$ to itself. Now let $S$ be any set.  If $a$
is any element of $S$, and if $K_a :S \to S$ is the constantly
$a$-valued morphism on $S$, then instantiating the naturality square
implied by the above equality gives that $G^\set_S \circ K_a = K_a
\circ G^\set_S$, i.e., $G^\set_S \, a = a$, i.e., $G^\set_S = \id_S$.
Putting these two cases together we have that for every $S : \set$,
$G^\set_S = \id_S$, i.e., $G^\set$ is the identity natural
transformation for the identity functor on $\set$. So every closed
term $g$ of closed type $\Nat^\alpha\alpha\,\alpha$ always denotes the
identity natural transformation for the identity functor on $\set$,
i.e., every closed term $g$ of type $\Nat^\alpha\alpha\,\alpha$
denotes the polymorphic identity function.

@@

As in~\cite{jg10}, replacing $\onet$ with any type $\emptyset;\alpha
\vdash C$ generalizes Theorem~\ref{thm:short-cut-nested} to deliver a
more general free theorem whose conclusion is $\mathit{fold}_{H}\, B
\; \circ \; G\; \mu H \; \mathit{in}_{H} = G
\,\setsem{\emptyset;\alpha \vdash K}\, B$.

@@

\noindent
Of course, we could have performed this replacement from the outset of
the present paper. But since using the above rule instead of the one
currently in Definition~\ref{def:wftypes} only makes the notation of
our calculus heavier without increasing its expressivity, we have
chosen not to.

@@

\noindent
respectively. Note that both of these rules make essential use of our
generalized $\Nat$-types. If $\Phi$ were required to be empty then we
would not be able to express, e.g., the types
\[\Nat^{\phi, \gamma,  \delta} (\phi \gamma \times
\phi \delta) \; (\phi (\gamma \times \phi))\] and
\[\Nat^{\phi, \beta}\,(\Lan^{\gamma,\delta}_{\gamma \times
  \delta} \,(\phi \gamma \times \phi \delta))\beta \; (\phi \beta)\]
associated with $\mathtt{Seq}$'s data constructor $\mathtt{spair}$.
